---
title: "Water Insecurity and Plumbing Access in U.S. Counties (2022-2023)"
subtitle: "Proposal"
author: "Adaria Blackwell and Jelani Butler"
 - name: "Blackwell-Butler"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Project description"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false

data_2022 <- read_csv("data/2025-01-28/water_insecurity_2022.csv")
data_2023 <- read_csv("data/2025-01-28/water_insecurity_2023.csv")



# Add a Year column to each if it’s not already there
data_2022 <- data_2022 |> mutate(Year = 2022)
data_2023 <- data_2023 |> mutate(Year = 2023)

# Combine them
water_data <- bind_rows(data_2022, data_2023)
```

### Dataset Description

This dataset contains U.S. county-level data on indoor plumbing access from the **American Community Survey (ACS)**, accessed using the `tidycensus` package. It captures social vulnerability indicators related to water insecurity for the years **2022 and 2023**.

Each row represents a single county in a specific year, and includes the following variables:

-   `GEOID`: Geographic identifier (FIPS code)\
-   `Name`: County and state\
-   `Year`: Year of the data (2022 or 2023)\
-   `Total_Pop`: Total population in the county\
-   `Plumbing`: Number of households with complete indoor plumbing\
-   `Percent_Lacking_Plumbing`: Percentage of households without complete plumbing\
-   `geometry`: Spatial polygon geometry for mapping

The dataset contains `r nrow(water_data)` observations from `r length(unique(water_data$Name))` unique counties. Data is provided for `r length(unique(water_data$Year))` years: `r paste(sort(unique(water_data$Year)), collapse = " and ")`.

This spatial dataset supports both statistical and geospatial analysis of plumbing access across the U.S., with built-in support for mapping using the `geometry` column.

I was curious to explore how access to something as essential as indoor plumbing varies across the U.S. It seemed like a challenge—and a unique opportunity—to better understand how plumbing access might be influenced by or influence population trends, and whether any geographic patterns stand out.

## Questions/Analysis plan

Question1: How has access to complete indoor plumbing changed from 2022 to 2023, and is this change associated with population size across U.S. counties.

-   Variables: Year, Percent_Lacking_Plumbing, and Total_Pop

-   Plan:

    -   Compare summary statistics and distributions across years.

    -   Use scatterplots or regression to examine how population size relates to changes in plumbing insecurity.

    -   Possibly calculate change scores (2023 % - 2022 %) if counties appear in both years.

Question2: Do counties with higher plumbing insecurity tend to cluster geographically, and does this clustering differ between 2022 and 2023?

-   Variables: Percent_Lacking_Plumbing, Name and GEOID (for location), and year

-   Plan:

    -   Group counties into "high", "moderate", and "low" plumbing insecurity.

    -   Create maps to visualize regional disparities and compare maps across years to identify shifts.

# Code

## Question 1:

```{r}
library(tidyverse)
library(sf)

# (Assume water_data has been loaded and GEOID fixed/renamed as needed)

# Reshape the data so that you have separate columns for 2022 and 2023 values:
change_data <- water_data %>%
  select(geoid, name, year, total_pop, percent_lacking_plumbing) %>%
  pivot_wider(names_from = year, values_from = c(total_pop, percent_lacking_plumbing)) %>%
  # Compute changes:
  mutate(
    change_plumbing = percent_lacking_plumbing_2023 - percent_lacking_plumbing_2022,
    change_population = total_pop_2023 - total_pop_2022
  )


ggplot(change_data, aes(x = change_population, y = change_plumbing)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  coord_cartesian(xlim = c(-100, 100)) +  # Zoom in on the x-axis
  labs(
  title = "Change in Plumbing Access vs. Change in Population (2022–2023)",
  subtitle = "Zoomed to ±100 population change",
  x = "Change in Population (2023 - 2022)",
  y = "Change in % Lacking Plumbing (2023 - 2022)"
) +
  theme_minimal()

```

```{r}
ggplot(change_data, aes(x = change_plumbing)) +
  geom_histogram(binwidth = 0.01, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Change in Plumbing Access (2022–2023)",
       x = "Change in % Lacking Plumbing", y = "Number of Counties") +
  theme_minimal()
```

## Question 2:

```{r}
  
AZwater22<-data_2022 %>%
    filter(geoid %in% c("04001", "04013", "04015", "04005", "04025", 
              "04021", "04027", "04017", "04003", "04019"))
              
glimpse(AZwater22)

AZwater23<-data_2023 %>%
  filter(geoid %in% c("04001", "04013", "04015", "04005", "04025", 
                      "04021", "04027", "04017", "04003", "04019"))

# now convert gemomrety from a text string to a gemoetry attribute

AZwater22geo<- AZwater22%>%
  mutate(
    geometry = purrr::map(geometry, \(geo) {
      eval(parse(text = geo))
    } )
  )

glimpse(AZwater22geo)
    
AZwater23geo<- AZwater23%>%
  mutate(
    geometry = purrr::map(geometry, \(geo) {
      eval(parse(text = geo))
    } )
  )
glimpse(AZwater23geo)

# Time to create the AZ map 

az_counties <- counties(cb = TRUE) |>
  st_transform("EPSG:5070") |>
  janitor::clean_names() |>
  filter(state_name == "Arizona")

#Join the data to make sure that both data sets are together

az22_map <- az_counties |>
  left_join(AZwater22geo, by = "geoid")

az23_map <- az_counties |>
  left_join(AZwater23geo, by = "geoid")

az23_map[c("name.x", "percent_lacking_plumbing")]
```

```{r}
#we are going to have to bin the data in order to make sure that we have a proper #ranges for percentages \`\`\`

az22bin <- az22_map |> mutate(plumbing_bins = cut( percent_lacking_plumbing, breaks = c(0, 1, 2, 3, 4, 5), labels = c("0-1%", "1-2%", "2-3%", "3-4%", "4-5%"), include.lowest = TRUE ))

glimpse(az22bin)

az23bin <- az23_map |> mutate(plumbing_bins = cut( percent_lacking_plumbing, breaks = c(0, 1, 2, 3, 4, 5), labels = c("0-1%", "1-2%", "2-3%", "3-4%", "4-5%"), include.lowest = TRUE ))

glimpse(az23bin)

#time to make the map after making the bins work

ggplot(data = az22bin) + geom_sf(aes(fill = plumbing_bins)) + geom_sf_text(aes(label = name.x), size = 3, color = "black") + scale_fill_manual( name = "Lacking Plumbing in 2022 (%)", values = c( "0-1%" = "#fee5d9", "1-2%" = "#fcae91", "2-3%" = "#fb6a4a", "3-4%" = "#de2d26", "4-5%" = "#a50f15"
),
na.value = "white"

) +

labs( x = "Longitude", y = "Latitude", title = "Percentage of Households Lacking Plumbing in Arizona (2022)" ) + theme_minimal()


```

```{r}
ggplot(data = az23bin) + geom_sf(aes(fill = plumbing_bins)) + geom_sf_text(aes(label = name.x), size = 3, color = "black") + scale_fill_manual( name = NULL, values = c( "0-1%" = "#fee5d9", "1-2%" = "#fcae91", "2-3%" = "#fb6a4a", "3-4%" = "#de2d26", "4-5%" = "#a50f15"
),
na.value = "white"
) +

labs( x = "Longitude", y = "Latitude", title = "Percentage of Households Lacking\nPlumbing in Arizona (2023)", caption = "SOURCE: Tidytuesday" ) + theme_minimal()
```

```{r}

```
