---
title: "Water Insecurity and Plumbing Access in U.S. Counties (2022-2023)"
subtitle: "Proposal"
author: "Adaria Blackwell and Jelani Butler"
 - name: "Team Name"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "Project description"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false

data_2022 <- read_csv("data/2025-01-28/water_insecurity_2022.csv")
data_2023 <- read_csv("data/2025-01-28/water_insecurity_2023.csv")



# Add a Year column to each if it’s not already there
data_2022 <- data_2022 |> mutate(Year = 2022)
data_2023 <- data_2023 |> mutate(Year = 2023)

# Combine them
water_data <- bind_rows(data_2022, data_2023)
```

### Dataset Description

This dataset contains U.S. county-level data on indoor plumbing access from the **American Community Survey (ACS)**, accessed using the `tidycensus` package. It captures social vulnerability indicators related to water insecurity for the years **2022 and 2023**.

Each row represents a single county in a specific year, and includes the following variables:

-   `GEOID`: Geographic identifier (FIPS code)\
-   `Name`: County and state\
-   `Year`: Year of the data (2022 or 2023)\
-   `Total_Pop`: Total population in the county\
-   `Plumbing`: Number of households with complete indoor plumbing\
-   `Percent_Lacking_Plumbing`: Percentage of households without complete plumbing\
-   `geometry`: Spatial polygon geometry for mapping

The dataset contains `r nrow(water_data)` observations from `r length(unique(water_data$Name))` unique counties. Data is provided for `r length(unique(water_data$Year))` years: `r paste(sort(unique(water_data$Year)), collapse = " and ")`.

This spatial dataset supports both statistical and geospatial analysis of plumbing access across the U.S., with built-in support for mapping using the `geometry` column.

I was curious to explore how access to something as essential as indoor plumbing varies across the U.S. It seemed like a challenge—and a unique opportunity—to better understand how plumbing access might be influenced by or influence population trends, and whether any geographic patterns stand out.

## Questions/Analysis plan

Question1: How has access to complete indoor plumbing changed from 2022 to 2023, and is this change associated with population size across U.S. counties.

-   Variables: Year, Percent_Lacking_Plumbing, and Total_Pop

-   Plan:

    -   Compare summary statistics and distributions across years.

    -   Use scatterplots or regression to examine how population size relates to changes in plumbing insecurity.

    -   Possibly calculate change scores (2023 % - 2022 %) if counties appear in both years.

Question2: Do counties with higher plumbing insecurity tend to cluster geographically, and does this clustering differ between 2022 and 2023?

-   Variables: Percent_Lacking_Plumbing, Name and GEOID (for location), and year

-   Plan:

    -   Group counties into "high", "moderate", and "low" plumbing insecurity.

    -   Create maps to visualize regional disparities and compare maps across years to identify shifts.
